|

### 2025-01-11 paper/manuscript.Rmd
SV edits

Studies on political change at the individual level---change in one's beliefs and preferences---face a challenging problem: while our methods typically provide estimates for change and stability at the group level, our goal is often to understand these processes at the individual level.

This *group-to-person generalizability problem*, the claim that group-level findings may not generalize to each person [@mcmanus2023], has important implications for what we can and can't say about political change across the life-course. While the previous research quantified change with various methods---ranging from assessing the time order of longitudinal observations [@kiley2020; @vaisey2020] to multilevel decomposition of over-time variance [@lersch2023]---these studies have fallen short of disentangling whether they find a small amount of change in a large number of people, or a large amount of change in a small number of people. In other words, while social science researchers have become widely sophisticated in evaluating the average levels of change over time in a population, they have mostly overlooked the fact that these average changes are produced by a combination of various compositional processes.

This issue arises in a variety of ways in empirical practice, particularly when we want to theorize the mechanisms of political belief change. For example, the General Social Survey (GSS) shows that roughly 35% of the American adults in 2006 believed that "marijuana should be legalized." In less than 10 years, this support had increased to nearly 60%. As seen in Figure 1, this change occurred both *cross-sectionally*, where average support increased with time, and *longitudinally*, where at least some segments of the American population changed their minds on this policy over these periods.

@@

While the cross-sectional processes have received strong scholarly attention in the literature [@vaisey2016; @restrepoochoa2024; @bartels2014], questions of the second sort remained largely unexplored. However, given that theories of political ideology involve theoretical expectations about individual-level outcomes, the fact that we lack information about whether our models capture person-level empirical processes is detrimental to understanding adult belief change in the political context. Therefore, our article aims to address this gap by examining whether we can identify (a) *how many people changed*, (b) *how much they changed*, and (c) *who changed*, based on observed survey response trajectories of political beliefs in empirical data.

Using simulations and empirical case studies, we show that it is indeed possible to provide plausible scenarios as to which generative models are more likely to have produced empirical observations, even though a reliable identification of *who* changed occurs only under very narrow conditions. In doing this, we build on the "Approximate Bayesian Computation" (ABC) literature [@beaumont2010] and provide a simple grid-search procedure to help social researchers in evaluating the extent to which their target estimates might be generalized to the population of their studies. Using this procedure, we show that it is possible to approximate to the underlying data generation processes for *rate of change*, *strength of change*, and *direction of change* in a panel data. We provide an R package, `gridsearch`, to help researchers in implementing this procedure in their own work.

We emphasize a striking heterogeneity in these underlying processes even for a seemingly straightforward case like the marijuana legalization in the GSS, showing that theoretically opposite claims---e.g., large swaths of the population increasing their support for the policy, as opposed to a small group of people completely switching their positions---may very well account for the same data. This finding is particularly important in the longitudinal panel context, which, for most applications, represents the gold standard in evaluating change in observational social research. As such, we claim that when the meaning of "political change" is elusive, and depends too much on how we define change in the first place, our current empirical strategies could potentially mislead rather than inform, especially if we overlook the underlying generative mechanisms.

We start with a general overview of the "latent variable model" of beliefs, the problems of reliability and resolution in survey measurement, and the compositional sources of aggregate change. We then introduce our grid-search procedure in the context of three-wave panel data with binary outcomes. Next, we relax this assumption and investigate how linear trajectories at the individual level can assist us in contexts with more than three waves. In the final section, we analyze whether this procedure helps us in classification tasks, particularly in evaluating *who* changed. We conclude by discussing the promises and pitfalls of panel data for understanding individual political change.

@@

The latent-variable approach reveals a fundamental problem in survey research: one's true score $Y_i$ is almost always realized with a certain degree of measurement error. Call this realization the *observed score* $y_i$. We often assume that $y_i$ represents the expected value of hypothetical draws from $Y_i$, with an $\epsilon$ term reflecting the measurement error in this process---i.e., $y_i = Y_i + \epsilon_i$. This error might result from various factors that introduce biases to how we extract information from individuals, given that problems like question construction and interviewer effects might contaminate how individuals reveal their true preferences [@alwin1991]. The inherent stochasticity in this measurement process implies that each draw from $Y_i$ might at least slightly differ from the true score, and this error term is difficult to circumvent with one measure, which is often the case in public opinion surveys.[^scales] In this sense, there is often a degree of systematic gap between $Y_i$ and $y_i$.

@@

In the panel context, where we observe the same people multiple times, this measurement problem has substantive implications: if true scores are imperfectly realized and we only have surface access to a person's observed survey responses, how can we know if they *actually* changed between time $t$ and $t + 1$, or whether the observed change is merely due to measurement error?

Scholars in public opinion research have attempted to address this issue in multiple ways, using structural equation models that explicitly quantify the measurement error component [@achen1975], strategies that integrate longitudinal surveys to exploit the information in the ordering of responses [@kiley2020], and various multilevel mixed-effects models that shrink responses based on population estimates to derive reliable trends [@lersch2023; @hout2016]. To some extent, all these strategies were fruitful to offset the problems associated with measurement error.[^stability] That said, while these empirical strategies were largely successful in accounting for the amount of change *in the population*, they were rather silent about changes *at the level of individuals*.

@@

Suppose we have a panel data, where individuals $i = 1, ..., N$ are observed over two time periods $t_0$ and $t_1$. Let $y_{it}$ represent an observed survey response. To see how much observed change there is at the individual level, we can simply define a change quantity, $\delta y_{i, t_0 \to t_1} = y_{it_1} - y_{it_0}$, the average of which, $\bar{y}$, represents the mean change in the dataset from $t_0$ to $t_1$.

This quantity is a natural candidate for describing aggregate belief change, but, under the hood, it summarizes three compositional processes in one metric. Let us define these processes intuitively. The first component is $R$, representing the proportion of individuals in our dataset with $\delta y_{i, t_0 \to t_1} \neq 0$, i.e., the proportion of individuals who changed in this period. Call this *the rate of change*. The second component is $S$, representing the "average" magnitude of change among those who changed.[^heterogeneity] Call this *the strength of change*. The third component is $D$, representing the net directionality of change, i.e., the difference in proportions of positive and negative changers in our observations. Call this *the direction of change*. If we combine these three components, we can represent $\bar{y}$ simply as

$$
\bar{y} = R \times S \times D
$$

[^heterogeneity]: Of course, we might further decompose this average strength to a distribution of heterogeneous change trajectories across individuals. For analytical and expository clarity, we bracket this issue from the scope of this article.

@@

For instance, if we want to evaluate whether a society became more pro-redistribution over time, we need to note that an increase in pro-redistributive preferences can occur via all or most people reporting higher preference for redistribution, some people switching their positions altogether while some others remain completely stable, or everyone changing in one or the other direction, and the net score we observe is simply the difference in their overall trajectories. The same reasoning applies to the GSS's marijuana legalization policy that we have discussed before. Most discussions in political belief change---confidence in legislative groups or financial institutions, beliefs about the economy, or ethnic attitudes---have the exact same question about their data generation processes. Simply put, the findings about change at the population level may simply result from a variety of conflicting, though observationally equivalent, data generation processes.

We can now complicate this composition question by noting, once again, the survey measurement problem of mapping one's true scores to their observed scores, which often fails in at least two ways: (1) our observed measures might have low *reliability*, such that a hypothetical set of repeated draws may significantly differ from one's true score, due to, e.g., poor question construction or individual idiosyncrasies, and (2) question response categories may be imperfect: the *resolution* of response categories may fail to distinguish consequential differences in true scores. Therefore, the average change captured in $\bar{y}$ may represent a biased or imperfect estimate of change to begin with.

@@

We pose these questions in a fixed-effects longitudinal panel data context [@allison2009; @vaisey2017], with varying individual trajectories across time periods [@rüttenauer2020]. This allows us to shift the focus from *aggregate change* to *individual change*, given that our theoretical interest is to understand how people change their political beliefs over the life-course.

@@

We build on the Approximate Bayesian Computation (ABC) literature [@beaumont2010] and propose a grid-search procedure to answer our research questions. The ABC approach involves simulating parameter values given the data and evaluating the distribution of these parameters that potentially generated our observations. We build a rejection sampler algorithm that

@@

3) Compares the simulated trajectories to the observed trajectories, calculating a "distance" score to quantify the extent to which specific simulated values reproduce the observed values;

>>>>

Studies on political change at the individual level---change in one's beliefs and preferences---face a challenging problem: while our methods typically provide estimates for change and stability at the population level, our goal is often to understand these processes at the individual level.

This *group-to-person generalizability problem*, the claim that group-level findings may not generalize to each person [@mcmanus2023], has important implications for what we can and can't say about political change across the life-course. While the previous research quantified change with various methods---ranging from assessing the time order of longitudinal observations [@kiley2020; @vaisey2020] to multilevel decomposition of over-time variance [@lersch2023]---these studies have been unable to disentangle whether they find a small amount of change in a large number of people, or a large amount of change in a small number of people. In other words, while social science researchers have become sophisticated in evaluating the average levels of change over time in a population, they have mostly overlooked the fact that these average changes can be produced by different processes.

This issue arises in many ways in empirical research, particularly when we want to theorize the mechanisms of political belief change. For example, the General Social Survey (GSS) shows that roughly 35% of the American adults in 2006 believed that "marijuana should be legalized." In less than 10 years, this support had increased to nearly 60%. As seen in Figure 1, this change occurred both *cross-sectionally*, where average support increased with time, and *longitudinally*, where at least some segments of the American population changed their minds on this policy over these periods.

@@

While the cross-sectional processes have received strong scholarly attention in the literature [@vaisey2016; @restrepoochoa2024; @bartels2014], questions of the second sort remained largely unexplored. However, given that theories of political ideology involve theoretical expectations about individual-level outcomes, the fact that we lack information about whether our models capture person-level empirical processes is detrimental to understanding adult political belief change. Therefore, our article aims to address this gap by examining whether we can identify (a) *how many people changed*, (b) *how much they changed*, and (c) *who changed*, based on observed survey response trajectories of political beliefs in empirical data.

Using simulations and empirical case studies, we show that it is indeed possible to use panel data to provide plausible estimates of which generative models are more likely to have produced particular empirical observations. We also find, however, that a reliable identification of *who* changed occurs only under very narrow conditions. In this sense, the modeling individual change problem is very much like the age-period-cohort identification problem that arises in repeated cross-sectional data [CITATION TO FOSSE]. In both cases, the problem cannot be "solved" by any particular methodological trick. Thus, our objective is not to provide a methodological panacea to eliminate the problem, but rather to provide a concrete approach that can help us understand the issue more clearly and manage it more effectively.

This approach builds on the "Approximate Bayesian Computation" (ABC) literature [@beaumont2010] and provides a simple grid-search procedure to help researchers evaluate the extent to which their target estimates might be generalized to a study population. Using this procedure, we show that it is possible to approximate to the underlying data generation processes for *rate of change*, *strength of change*, and *direction of change* in a panel dataset. We provide an R package, `gridsearch`, to help researchers implement this procedure in their own work.

We emphasize the existence of striking heterogeneity in these underlying processes even for a seemingly straightforward case like the marijuana legalization in the GSS; that is, we show that theoretically opposite claims---e.g., large swaths of the population increasing their support for the policy, as opposed to a small group of people completely switching their positions---may very well account for the same data. This finding is particularly important in the longitudinal panel context, which, for most applications, represents the gold standard in evaluating change in observational research. For this reason, we argue that when the meaning of "political change" is not clearly specified, our current empirical strategies could potentially mislead rather than inform, especially if we overlook the underlying generative mechanisms.

We begin with a general overview of the "latent variable model" of beliefs, the problems of reliability and resolution in survey measurement, and the compositional sources of aggregate change. We then introduce our grid-search procedure in the context of three-wave panel data with binary outcomes. Next, we relax this assumption and investigate how linear trajectories at the individual level can assist us in contexts with more than three waves. In the final section, we analyze whether this procedure helps us in classification tasks, particularly in evaluating *who* changed. We conclude by discussing the promises and pitfalls of panel data for understanding individual political change.

@@

The latent-variable approach acknowledges a fundamental problem in survey research: one's true score $Y_i$ is almost always realized with a certain degree of measurement error. Call this realization the *observed score*, $y_i$. We often assume that $y_i$ represents the expected value of hypothetical draws from $Y_i$, with an $\epsilon$ term reflecting the measurement error in this process---i.e., $y_i = Y_i + \epsilon_i$. This error might result from various factors that introduce biases in how we extract information from individuals, given that problems like question construction and interviewer effects might influence how individuals reveal their true preferences [@alwin1991]. The inherent randomness in the measurement process implies that each draw from $Y_i$ might at least slightly differ from the true score, and this error term is difficult to circumvent with a one-item measure, which is often the case in large public opinion surveys.[^scales] In this sense, there is often a degree of systematic gap between $Y_i$ and $y_i$.

@@

In the panel context, where we observe the same people multiple times, this measurement problem has substantive implications: if true scores are imperfectly realized and we only have access to a person's observed survey responses, how can we know if they *actually* changed between time $t$ and $t + 1$, or whether the observed change is merely due to measurement error?

Scholars in public opinion research have attempted to address this issue in multiple ways, using structural equation models that explicitly quantify the measurement error component [@achen1975], strategies that integrate longitudinal surveys to exploit the information in the ordering of responses [@kiley2020], and various multilevel mixed-effects models that shrink responses based on population estimates to derive reliable trends [@lersch2023; @hout2016]. To some extent, all these strategies have been useful for offseting the problems associated with measurement error.[^stability] That said, while these empirical strategies have been largely successful in accounting for the amount of change *in the population*, they were rather silent about changes *at the level of individuals*.

@@

Suppose we have a panel data, where individuals $i = 1, ..., N$ are observed over two time periods $t_0$ and $t_1$. Let $y_{it}$ represent an observed survey response. To see how much observed change there is at the individual level, we can simply define a change quantity, $\delta y_{i, t_0 \to t_1} = y_{it_1} - y_{it_0}$, the average of which, $\bar{\delta_y}$, represents the mean change in the dataset from $t_0$ to $t_1$.

This quantity is a natural candidate for describing aggregate belief change, but, under the hood, it summarizes three compositional processes in one metric. Let us define these processes intuitively. The first component is $R$, or the *rate of change*. This represents the proportion of individuals in our dataset with $\delta y_{i, t_0 \to t_1} \neq 0$, i.e., the proportion of individuals who changed in this period. The second component is $S$, the *strength of change*. This represents the "average" magnitude of change among those who changed.[^heterogeneity] The third component is $D$, the *direction of change*. This represents the net directionality of change, i.e., the difference in proportions of positive and negative changers in our observations. If we combine these three components, we can represent $\bar{y}$ simply as

$$
\bar{\delta_y} = R \times S \times D
$$

[^heterogeneity]: Of course, we might further decompose this average strength to a distribution of heterogeneous change trajectories across individuals. For analytical and expository clarity, we bracket this issue from in this article.

@@

For instance, if we want to evaluate how a society became more pro-redistribution over time, we need to note that an increase in pro-redistributive preferences could occur via all or most people reporting higher preference for redistribution, some people switching their positions altogether while some others remain completely stable, or everyone changing in one or the other direction, and the net score we observe is simply the difference in their overall trajectories. The same reasoning applies to the GSS's marijuana legalization policy we discussed above. Most discussions in political belief change---confidence in legislative groups or financial institutions, beliefs about the economy, or ethnic attitudes---have the exact same question about their data generation processes. Simply put, the findings about individual change at the population level may simply result from a variety of incompatible, though observationally equivalent, data generation processes.[^APC]

[^APC]: Again, this is a very similar problem to the age-period-cohort identification problem we mentioned above.

As if this weren't complicated enough, we must complicate the issue further by, once again, recognizing the measurement problem of mapping true scores to observed scores. This process often fails in at least two ways: (1) our observed measures might have low *reliability*, such that a hypothetical set of repeated draws may significantly differ from one's true score, and (2) question response categories may be imperfect: the *resolution* of response categories may fail to distinguish consequential differences in true scores. Therefore, the average change captured in $\bar{y}$ may represent a biased or imperfect estimate of change to begin with.

@@

We pose this question in a fixed-effects longitudinal panel data context [@allison2009; @vaisey2017], with varying individual trajectories across time periods [@rüttenauer2020]. This allows us to shift the focus from *aggregate change* to *individual change*, given that our theoretical interest is to understand how people change their political beliefs over the life-course.

@@

We build on the Approximate Bayesian Computation (ABC) literature [@beaumont2010] and propose a grid-search procedure to answer this research question. The ABC approach involves simulating parameter values given the data and evaluating the distribution of these parameters that potentially generated our observations. We build a rejection sampler algorithm that

@@

3) Compares the distribution of simulated trajectories to the distribution of observed trajectories, calculating a "distance" score to quantify the extent to which specific simulated values reproduce the observed values;
